---
title: 'Microbiome Rubiceli '
author: "Luigui Gallardo Becerra, Dr. Adrián Ochoa Leyva"
date: "8/12/2021"
output:
  html_document: default
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(engine.opts = list(bash = "-l"))
knitr::opts_knit$set(root.dir = '/Users/luigui/Documents/rubiceli_microbiome_16S')

# Packages installation
library(colorRamps)
library(ggplot2)
library(reshape)
library(scales)
library(jcolors)
library(grid)
library(RColorBrewer)
library(dplyr)
library(ggpubr)
library(ggordiplots)

```

## Taxonomy, only **Initial feces** phylum level

```{r, echo=FALSE, warning=FALSE, include=FALSE}
# Input table
table <- read.table(file = "01_taxonomy/01_tax_sample/formatted_table_L2.txt",
      header = TRUE)

# Reshape table
melted_table <- melt(table, id.vars = c("SampleID", "Description"))

# Rename headers
melted_table <- melted_table %>% 
  rename(sample = SampleID,
    group = Description,
    taxa = variable)

# Filter only taxa
filt_table <- melted_table %>% filter(grepl("k__", taxa))

filt_table$taxa <- gsub('k__Bacteria.p__', '', filt_table$taxa)

filt_table$value <- as.numeric(as.character(filt_table$value))

filt_table <- subset(filt_table, taxa != "")

# Filter only "Initial" samples
filt_table <- filt_table %>% filter(grepl("Inicial", group))

# Change name of group
filt_table$group <- recode_factor(filt_table$group, Inicial = "Initial feces")

# Colors
colourCount = length(unique(filt_table$taxa))
getPalette = colorRampPalette(brewer.pal(colourCount, "Set3"))

# Plot
barplot_tax <- ggplot(filt_table, aes(x = sample,
    y = value,
    fill = taxa)) +
geom_bar(position="fill", stat="identity") +
facet_grid(~group,
    scales = "free") +
labs(x = "",
    y = "Relative abundance",
    fill = toupper("phylum")) +
scale_x_discrete(expand = c(0, 0)) +
scale_y_continuous(limits = c(0, 1),
    expand = c(0, 0)) +
scale_fill_manual(values = getPalette(colourCount)) +
theme_test(base_size = 8,
           base_family = "Arial") +
theme(legend.title = element_text(hjust = 1),
      legend.key.height = unit(1, "line"),
      legend.key.width = unit(1, "line")) +
guides(fill = guide_legend(ncol = 1,
    byrow = FALSE))

# Creation of .svg
ggsave("01_taxonomy/01_tax_sample/phylum_initial-feces.svg",
       width = 3,
       height = 4)
print(barplot_tax)
dev.off()
```

![phylum_initial-feces](01_taxonomy/01_tax_sample/phylum_initial-feces.svg)

## Alpha diversity, ND vs HFD

```{r, echo=FALSE, warning=FALSE, include=FALSE}
# Input table
alpha_data <- read.table(file = "02_alpha_diversity/alpha_diversity_e24251_r10000.txt",
                         header = TRUE,
                         sep = "\t")

# Filter groups DN and DAG
filt_alpha_data <- alpha_data %>% filter(grepl('DN|DAG',
                                               Grupo_Experimental))

filt_alpha_data <- filt_alpha_data %>% filter(!grepl('_A',
                                               Grupo_Experimental))

# Change name of groups
filt_alpha_data$Grupo_Experimental <- recode_factor(filt_alpha_data$Grupo_Experimental,
                                  DAG = "HFD",
                                  DN = "ND")

filt_alpha_data <- filt_alpha_data %>% rename(group = "Grupo_Experimental")

filt_alpha_data <- filt_alpha_data %>% mutate(group = factor(group,
                                                             levels=c("ND",
                                                                      "HFD")))
# Define colors and shapes
colors <- c("#6600cc", # purple
            "#800000") # brown

shapes <- c(8, # *
            3) # +

# P-value comparisons 
p_value <- list(c("ND",
                  "HFD"))

# Observed otus boxplot
observed_otus_boxplot <- ggplot(filt_alpha_data,
                                aes(x = group,
                                    y = Observed.OTUs)) +
geom_boxplot(aes(color = group),
             outlier.shape = NA) +
geom_point(aes(color = group),
    size = 2,
    shape = 21,
    position = position_jitterdodge()) +
theme_test(base_size = 8,
           base_family = "Arial") +
labs(x = "",
    y = "Observed OTUs") +
scale_color_manual(values = colors) +
stat_compare_means(comparisons = p_value,
    method = "wilcox.test") +
theme(legend.position = "none")

# Creation of .svg
ggsave("02_alpha_diversity/observed_nd_vs_hdf.svg",
       width = 4,
       height = 4)
print(observed_otus_boxplot)
dev.off()

```

![observed_nd_vs_hdf](02_alpha_diversity/observed_nd_vs_hdf.svg)

```{r, echo=FALSE, warning=FALSE, include=FALSE}
# Shannon boxplot
shannon_boxplot <- ggplot(filt_alpha_data,
                                aes(x = group,
                                    y = Shannon.Index)) +
geom_boxplot(aes(color = group),
             outlier.shape = NA) +
geom_point(aes(color = group),
    size = 2,
    shape = 21,
    position = position_jitterdodge()) +
theme_test(base_size = 8,
           base_family = "Arial") +
labs(x = "",
    y = "Shannon Index") +
scale_color_manual(values = colors) +
stat_compare_means(comparisons = p_value,
    method = "wilcox.test") +
theme(legend.position = "none")

# Creation of .svg
ggsave("02_alpha_diversity/shannon_nd_vs_hdf.svg",
       width = 4,
       height = 4)
print(shannon_boxplot)
dev.off()
```

![shannon_nd_vs_hdf](02_alpha_diversity/shannon_nd_vs_hdf.svg)

## Beta diverstity, Unweighted UniFrac ND vs HFD

```{r, echo=FALSE, warning=FALSE, include=FALSE}
# Input table and metadata
beta_data <- read.table(file = "03_beta_diversity/01_nd_vs_hfd//unweighted_unifrac_dm.txt",
                        sep="\t",
                        header=T,
                        row.names=1,
                        check.names=F)

metadata <- read.table(file = "metadata_allsamples.txt",
                       sep="\t",
                       header=T,
                       row.names=1,
                       check.names=F)

metadata <- metadata %>% filter(grepl("D", Grupo_Experimental)) %>% filter(!grepl("_", Grupo_Experimental))

# Change the name of groups
metadata$Grupo_Experimental <- recode_factor(metadata$Grupo_Experimental,
                                             DN = "ND",
                                             DAG = "HFD")

metadata <- metadata %>% mutate(Grupo_Experimental = factor(Grupo_Experimental,
                                                            levels=c(
                                                              "ND",
                                                              "HFD")))

# Define colors and shapes
colors <- c("#6600cc", # purple
            "#800000") # brown

shapes <- c(8, # *
            3) # +

# PCoA plot
ord <- cmdscale(beta_data,
                eig = T,
                k=ncol(beta_data)-1)

plt <- gg_ordiplot(ord,
                   groups = metadata$Grupo_Experimental,
                   conf = 0.95,
                   spiders = TRUE,
                   plot = FALSE,
                   ellipse = FALSE,
                   pt.size = 1)

plt <- plt$plot +
  geom_point(data = plt$df_ord,
             aes(x = x, y = y, color = Group, shape = Group),
             size = 2) +
scale_color_manual(values = colors) +
scale_shape_manual(values = shapes) +
labs(y = "PC2 (16.55%)",
     x = "PC1 (49.09%)",
    title = "ANOSIM=0.9630; p-value=0.002") +
geom_hline(yintercept=0,
    color = "gray") +
geom_vline(xintercept=0,
    color = "gray") +
theme_test(base_size = 8,
           base_family = "Arial") +
theme(legend.title = element_blank(),
    legend.position = "right",
    legend.background = element_blank(),
    plot.title = element_text(hjust = 0.5))


# Creation of .svg
ggsave("03_beta_diversity/01_nd_vs_hfd/unweighted_unifrac_nd_vs_hfd.svg",
       width = 4.5,
       height = 4)
print(plt)
dev.off()
```

![unweighted_unifrac_nd_vs_hfd](03_beta_diversity/01_nd_vs_hfd/unweighted_unifrac_nd_vs_hfd.svg)

## Taxonomy, 12 weeks ND vs HFD species level

```{r, echo=FALSE, warning=FALSE, include=FALSE}
# Input table
table <- read.table(file = "01_taxonomy/02_tax_group/formatted_table_L7_v2.txt",
      header = TRUE,
      sep = '\t',
      check.names = FALSE)

# Reshape table
melted_table <- melt(table, id.vars = c("group"))

# Rename headers
filt_table <- melted_table %>% 
  rename(taxa = variable)

# Rename groups
filt_table$group <- recode_factor(filt_table$group,
                                  DAG = "HFD",
                                  DN = "ND")

filt_table <- filt_table %>% mutate(group = factor(group,
                                                   levels=c("ND", "HFD")))
# Colors
colourCount = length(unique(filt_table$taxa))
getPalette = colorRampPalette(brewer.pal(colourCount, "Set3"))

# Plot
barplot_tax <- ggplot(filt_table, aes(x = group,
    y = value,
    fill = taxa)) +
geom_bar(position="stack", stat="identity") +
labs(x = "",
    y = "Relative abundance",
    fill = toupper("species")) +
scale_x_discrete(expand = c(0, 0)) +
scale_y_continuous(limits = c(0, 0.45),
   expand = c(0, 0)) +
scale_fill_manual(values = getPalette(colourCount)) +
theme_test(base_size = 8,
           base_family = "Arial") +
theme(legend.title = element_text(hjust = 1),
      legend.key.height = unit(1, "line"),
      legend.key.width = unit(1, "line")) +
guides(fill = guide_legend(ncol = 1,
    byrow = FALSE))

# Creation of .svg
ggsave("01_taxonomy/02_tax_group/species_12w-ND-HFD.svg",
       width = 3,
       height = 4)
print(barplot_tax)
dev.off()
```

![species_12w-ND-HFD](01_taxonomy/02_tax_group/species_12w-ND-HFD.svg)

## Alpha diversity, NE vs EE

```{r, echo=FALSE, warning=FALSE, include=FALSE}
# Input table
alpha_data <- read.table(file = "02_alpha_diversity/alpha_diversity_e24251_r10000.txt",
                         header = TRUE,
                         sep = "\t")

# Filter groups DN and DAG
filt_alpha_data <- alpha_data %>% filter(grepl('DN|DAG',
                                               Grupo_Experimental))

filt_alpha_data <- filt_alpha_data %>% filter(grepl('_A',
                                               Grupo_Experimental))

# Change name of groups
filt_alpha_data$Grupo_Experimental <- recode_factor(filt_alpha_data$Grupo_Experimental,
                                  DAG_AE = "HFD EE",
                                  DN_AE = "ND EE",
                                  DAG_AN = "HFD NE",
                                  DN_AN = "ND NE")

filt_alpha_data$Dieta <- recode_factor(filt_alpha_data$Dieta,
                                  DAG = "HFD",
                                  Normal = "ND")

filt_alpha_data <- filt_alpha_data %>% rename(group = "Grupo_Experimental",
                                              diet = "Dieta")

filt_alpha_data <- filt_alpha_data %>% mutate(diet = factor(diet,
                                                             levels=c("ND",
                                                                      "HFD")))

filt_alpha_data <- filt_alpha_data %>% mutate(group = factor(group,
                                                             levels=c("ND NE",
                                                                      "ND EE",
                                                                      "HFD NE",
                                                                      "HFD EE")))

# Define colors
colors <- c("#0000FF", # blue
            "#EDBD04", # yellow
            "#FB0207", # red
            "#000000") # black

# P-value comparisons 
p_value <- list(c("ND NE", "ND EE"),
                c("HFD NE", "HFD EE"))

# Observed otus boxplot
observed_otus_boxplot <- ggplot(filt_alpha_data,
                                aes(x = group,
                                    y = Observed.OTUs)) +
geom_boxplot(aes(color = group),
             outlier.shape = NA) +
geom_point(aes(color = group),
    size = 2,
    shape = 21,
    position = position_jitterdodge()) +
# facet_grid(~diet,
#     scales = "free") +
theme_test(base_size = 8,
           base_family = "Arial") +
labs(x = "",
    y = "Observed OTUs") +
scale_color_manual(values = colors) +
stat_compare_means(comparisons = p_value,
    method = "wilcox.test") +
theme(legend.position = "none")

# Creation of .svg
ggsave("02_alpha_diversity/observed_ee_vs_ne.svg",
       width = 4,
       height = 4)
print(observed_otus_boxplot)
dev.off()

```

![observed_ee_vs_ne](02_alpha_diversity/observed_ee_vs_ne.svg)

```{r, echo=FALSE, warning=FALSE, include=FALSE}
# Shannon boxplot
shannon_boxplot <- ggplot(filt_alpha_data,
                                aes(x = group,
                                    y = Shannon.Index)) +
geom_boxplot(aes(color = group),
             outlier.shape = NA) +
geom_point(aes(color = group),
    size = 2,
    shape = 21,
    position = position_jitterdodge()) +
# facet_grid(~diet,
#     scales = "free") +
theme_test(base_size = 8,
           base_family = "Arial") +
labs(x = "",
    y = "Shannon Index") +
scale_color_manual(values = colors) +
stat_compare_means(comparisons = p_value,
    method = "wilcox.test") +
theme(legend.position = "none")

# Creation of .svg
ggsave("02_alpha_diversity/shannon_ee_vs_ne.svg",
       width = 4,
       height = 4)
print(shannon_boxplot)
dev.off()
```

![shannon_ee_vs_ne](02_alpha_diversity/shannon_ee_vs_ne.svg)

## Beta diversity, Unweighted UniFrac all samples

```{r, echo=FALSE, warning=FALSE, include=FALSE}
# Input table and metadata
beta_data <- read.table(file = "03_beta_diversity/02_allgroups//unweighted_unifrac_dm.txt",
                 sep="\t",
                 header=T,
                 row.names=1,
                 check.names=F)

metadata <- read.table(file = "metadata_allsamples.txt",
                  sep="\t",
                  header=T,
                  row.names=1,
                  check.names=F)

# Change the name of groups
metadata$Grupo_Experimental <- recode_factor(metadata$Grupo_Experimental,
                                 Inicial = "Initial",
                                 DN = "ND",
                                 DAG = "HFD",
                                 DN_AN = "ND NE",
                                 DN_AE = "ND EE",
                                 DAG_AN = "HFD NE",
                                 DAG_AE = "HFD EE")

metadata <- metadata %>% mutate(Grupo_Experimental = factor(Grupo_Experimental,
                                                             levels=c(
                                                               "Initial",
                                                               "ND",
                                                               "HFD",
                                                               "ND NE",
                                                               "ND EE",
                                                               "HFD NE",
                                                               "HFD EE")))

# Define colors and shapes
colors <- c("#4d4d4d", # gray
            "#6600cc", # purple
            "#800000", # brown
            "#0000FF", # blue
            "#EDBD04", # yellow
            "#FB0207", # red
            "#000000") # black

shapes <- c(4, # x
            8, # *
            3, # +
            16, # circle
            17, # triangle
            15, # square
            18) # diamond

# PCoA plot
ord <- cmdscale(beta_data,
                eig = T,
                k=ncol(beta_data)-1)

plt <- gg_ordiplot(ord,
                   groups = metadata$Grupo_Experimental,
                   conf = 0.95,
                   spiders = TRUE,
                   plot = FALSE,
                   ellipse = FALSE,
                   pt.size = 1)

plt <- plt$plot +
geom_point(data = plt$df_ord,
             aes(x = x, y = y, color = Group, shape = Group),
    size = 2) +
scale_color_manual(values = colors) +
scale_shape_manual(values = shapes) +
labs(y = "PC2 (7.49%)",
     x = "PC1 (48.11%)") +
geom_hline(yintercept=0,
    color = "gray") +
geom_vline(xintercept=0,
    color = "gray") +
theme_test(base_size = 8,
           base_family = "Arial") +
theme(legend.title = element_blank(),
    legend.position = "right",
    legend.background = element_blank(),
    plot.title = element_text(hjust = 0.5))

# Creation of .svg
ggsave("03_beta_diversity/02_allgroups/unweighted_unifrac_allgroups.svg",
       width = 4.5,
       height = 4)
print(plt)
dev.off()
```

![unweighted_unifrac_allgroups](03_beta_diversity/02_allgroups/unweighted_unifrac_allgroups.svg)

## Beta diversity, Unweighted UniFrac HFD NE vs EE

```{r, echo=FALSE, warning=FALSE, include=FALSE}
# Input table and metadata
beta_data <- read.table(file = "03_beta_diversity/03_nd_vs_hfd_env/weighted_unifrac_dm.txt",
                        sep="\t",
                        header=T,
                        row.names=1,
                        check.names=F)

metadata <- read.table(file = "metadata_allsamples.txt",
                       sep="\t",
                       header=T,
                       row.names=1,
                       check.names=F)

metadata <- metadata %>% filter(grepl("DAG", Grupo_Experimental)) %>% filter(grepl("_", Grupo_Experimental))

# Change the name of groups
metadata$Grupo_Experimental <- recode_factor(metadata$Grupo_Experimental,
                                             DAG_AN = "HFD NE",
                                             DAG_AE = "HFD EE")

metadata <- metadata %>% mutate(Grupo_Experimental = factor(Grupo_Experimental,
                                                            levels=c(
                                                              "HFD NE",
                                                              "HFD EE")))

# Define colors and shapes
colors <- c("#FB0207", # red
            "#000000") # black

shapes <- c(15, # square
            18) # diamond

# PCoA plot
ord <- cmdscale(beta_data,
                eig = T,
                k=ncol(beta_data)-1)

plt <- gg_ordiplot(ord,
                   groups = metadata$Grupo_Experimental,
                   conf = 0.95,
                   spiders = TRUE,
                   plot = FALSE,
                   ellipse = FALSE,
                   pt.size = 1)

plt <- plt$plot +
  geom_point(data = plt$df_ord,
             aes(x = x, y = y, color = Group, shape = Group),
             size = 2) +
scale_color_manual(values = colors) +
scale_shape_manual(values = shapes) +
labs(y = "PC2 (15.24%)",
     x = "PC1 (27.90%)",
    title = "ANOSIM=0.6111; p-value=0.003") +
geom_hline(yintercept=0,
    color = "gray") +
geom_vline(xintercept=0,
    color = "gray") +
theme_test(base_size = 8,
           base_family = "Arial") +
theme(legend.title = element_blank(),
    legend.position = "right",
    legend.background = element_blank(),
    plot.title = element_text(hjust = 0.5))

# Creation of .svg
ggsave("03_beta_diversity/03_nd_vs_hfd_env/unweighted_unifrac_nd_vs_hfd_env.svg",
       width = 4.5,
       height = 4)
print(plt)
dev.off()
```

![unweighted_unifrac_allgroups](03_beta_diversity/03_nd_vs_hfd_env/unweighted_unifrac_nd_vs_hfd_env.svg)

